BlockTensorDecompositions.jl: A Unified Constrained Tensor Decomposition Julia Package

### 2025-02-04 paper/paper.qmd
first pass of sub-block descent section

@@

@@

The (complete) multi-mode product $\times_{1,\dots,n}=\times_{[n]}$ between a tensor $A\in\mathbb{R}^{I_1 \times \dots \times I_N}$ and tensor $B\in\mathbb{R}^{I_1\times\dots\times I_n}$, returns a tensor $(A \times_{[n]} B)\in\mathbb{R}^{I_{n+1} \times \dots \times I_N}$ with entries

$$
(A \times_{[n]} B)[i_{n+1},\dots,i_N] = \sum_{i_1, \dots ,i_n} A[i_1, \dots,i_n,i_{n+1},\dots,i_N]B[i_1, \dots ,i_n].
$$

@@

@@

The Hessian $\nabla^2 f:\mathbb{R}^{I_1\times \dots \times I_N}\to \mathbb{R}^{(I_1\times \dots \times I_N)^2}$ of a (second-differentiable) function $f:\mathbb{R}^{I_1\times \dots \times I_N}\to \mathbb{R}$ is the gradient of the gradient and is defined for a tensor $A\in\mathbb{R}^{I_1\times \dots \times I_N}$ entry-wise by

@@

@@

For matrices $M$, the (operator) 2-norm is defined as

$$
\left\lVert M \right\rVert_{\text{op}} = \left\lVert M \right\rVert_2 = \sup_{\lVert v \rVert_2 = 1} \left\lVert Mv \right\rVert_2 = \sigma_1(M)
$$

@@

For tensors $T$, the (operator) 2-norm needs to be defined in terms of how we treat them as function on other tensors. There is a canonical way to do this for vectors $x \to v^{\top}x$ and matrices $x \to Mx$, but not tensors. This is relevant to @sec-lipschitz-computation where the Lipschitz step-size is computed in terms of the Operator norm of the Hessian of our objective function. In the case of the Hessian tensor $\nabla^2 f(A)$ evaluated at $A$, we consider the function $X \to \nabla^2 f(A) \times_{[N]} X$. This gives us

$$
\left\lVert \nabla^2 f(A) \right\rVert_{\text{op}} = \sup_{\left\lVert X \right\rVert_F = 1} \left\lVert \nabla^2 f(A) \times_{[N]} X \right\rVert_F = \sigma_1(\text{mat}(\nabla^2 f(A)))
$$

where $\text{mat}:\mathbb{R}^{(I_1\times \dots \times I_N)^2}\to\mathbb{R}^{(I_1 \dots I_N)\times (I_1 \dots I_N)}$ takes a $2N$-ordered tensor and flattens it into a square $(I_1 \dots I_N)\times (I_1 \dots I_N)$ matrix.

@@

@@

Then

@@

$$
L = \left\lVert D \right\rVert _{\text{op}} = \sigma_1(\operatorname{mat}D).
$$

@@

For the family of Tucker decompositions, we can compute the Lipschitz constants efficiently similar to how we compute the gradient in @sec-gradient-computation.

::: callout-note
For the family of Tucker decompositions, the objective function $f$ is "block-quadratic" with respect to the factors. We saw in @eq-tucker-1-gradient-2 that the gradient is an affine map where $B$ is multiplied by the "slope" $A^\top A$ plus a shift of $-Y\times_1 A^\top$.

@@

TODO Explain how we get the following. General idea is to take partial Hessian of our quadratic function with respect to one block, and get that the operator norm of the hessian (as a map from $\mathbb{R}^{I_1\times \dots \times I_N}$ to itself) is the Lipschitz constant of the gradient. See Obsidian note "Gradients Hessians and Lipschitz Constants of a Tucker Decomposition".

```julia
function make_lipschitz(T::Tucker1, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    if n==0 # the core is the zeroth factor
        function lipschitz0(T::Tucker1; kwargs...)
            A = matrix_factor(T, 1)
            return opnorm(A'A)
        end
        return lipschitz0

    elseif n==1 # the matrix is the zeroth factor
        function lipschitz1(T::Tucker1; kwargs...)
            C = core(T)
            return opnorm(slicewise_dot(C, C))
        end
        return lipschitz1

    else
        error("No $(n)th factor in Tucker1")
    end
end

function make_lipschitz(T::Tucker, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    N = ndims(T)
    if n==0 # the core is the zeroth factor
        function lipschitz_core(T::AbstractTucker; kwargs...)
            #matrices = matrix_factors(T)
            #gram_matrices = map(A -> A'A, matrices)
            #return prod(opnorm.(gram_matrices))
            return prod(A -> opnorm(A'A), matrix_factors(T))
        end
        return lipschitz_core

    elseif n in 1:N # the matrix is the zeroth factor
        function lipschitz_matrix(T::AbstractTucker; kwargs...)
            matrices = matrix_factors(T)
            TExcludeAn = tuckerproduct(core(T), matrices; exclude=n)
            return opnorm(slicewise_dot(TExcludeAn, TExcludeAn; dims=n))
        end
        return lipschitz_matrix

    else
        error("No $(n)th factor in Tucker")
    end
end

function make_lipschitz(T::CPDecomposition, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    N = ndims(T)
    if n in 1:N # the matrix is the zeroth factor
        function lipschitz_matrix(T::AbstractTucker; kwargs...)
            matrices = matrix_factors(T)
            TExcludeAn = tuckerproduct(core(T), matrices; exclude=n) # TODO optimize this to avoid making the super diagonal core
            return opnorm(slicewise_dot(TExcludeAn, TExcludeAn; dims=n))
        end
        return lipschitz_matrix

@@

@@

In [@lee_algorithms_2000], the multiplicative update algorithm used to solve nonnegative matrix factorization problems is written in a way so that each row of the first factor

@@

@@

>>>>

::: {#def-outer-product}

@@

The $n$-slice product of a tensor with itself $X \cdot_n X$ should be thought of as a generalization of the Gram matrix $X^\top X$ since it considers the matrix generated by taking the dot product between every $n$th mode slice, just like how the Gram matrix considers the dot product between every pair of columns.

@@

The multi-mode product $\times_{1,\dots,n}=\times_{1:n}= \times_{[n]}$ between a tensor $A\in\mathbb{R}^{I_1 \times \dots \times I_{N}}$ and tensor $B\in\mathbb{R}^{I_1\times \dots\times I_{n}}$, returns a tensor $(A \times_{[n]} B)\in\mathbb{R}^{I_{n+1} \times \dots \times I_{N}}$ with entries

$$
(A \times_{[n]} B)[i_{n+1},\dots,i_{N}] = \sum_{i_1, \dots ,i_n} A[i_1, \dots,i_{n}, i_{n+1}, \dots,i_{N}]B[i_1,\dots,i_{n}].
$$

This product contracts the first $n$ indexes of $A$ with every index of $B$.

More generally, we can contract any number of indexes such as the last $n$ indexes of $A$ with every index of $B$ with $\times_{N-n+1,\dots,N}=\times_{(N-n+1):N}=\times_{[-n]}$,

$$
(A \times_{[-n]} B)[i_1, \dots,i_{n}] = \sum_{i_{n+1}, \dots,i_{N}} A[i_1, \dots,i_{N-n}, i_{N-n+1}, \dots,i_{N}]B[i_{N-n+1}, \dots,i_{N}],
$$

or specific indexes. For example, we would define $(A\times_{1,3,5}B)\in\mathbb{R}^{I_2\times I_4\times I_6}$ where $A\in\mathbb{R}^{I_1\times\dots\times I_6}$ and $B\in\mathbb{R}^{I_1\times I_3\times I_5}$ to be

$$
(A \times_{1,3,5} B)[i_2,i_4,i_6] = \sum_{i_1,i_3,i_5} A[i_1,i_2,i_3,i_4,i_5,i_6]B[i_1,i_3,i_5].
$$

When $A$ a *half-symmetric*^[For example, the Hessian of a scalar function (see @def-hessian).] tensor of order $2N$

$$
A[i_1, \dots,i_{N}, i_{N+1}, \dots,i_{2N}] = A[i_{N+1}, \dots,i_{2N},i_1, \dots,i_{N}],
$$ {#eq-half-symmetric}

we have

$$
A \times_{[N]} B = A \times_{[-N]} B
$$

for tensors $B$ of order $N$.

<!-- The half product $\times$ between a tensor $A\in\mathbb{R}^{I_1 \times \dots \times I_{2N}}$ and tensor $B\in\mathbb{R}^{I_2\times I_4\times\dots\times I_{2N}}$, returns a tensor $(A \times B)\in\mathbb{R}^{I_1 \times I_3 \times \dots \times I_{2N-1}}$ with entries

$$
(A \times B)[i_1,i_3,\dots,i_{2N-1}] = \sum_{i_1, \dots ,i_n} A[i_1, \dots,i_{2N}]B[i_2,i_4,\dots,i_{2N}].
$$

This product contracts every other index of $A$ with every index of $B$. -->

@@

::: {#def-gradient}

# Gradient

@@

::: {#def-hessian}

# Hessian

The Hessian $\nabla^2 f:\mathbb{R}^{I_1\times \dots \times I_N}\to \mathbb{R}^{(I_1\times \dots \times I_N)^2}$ of a second-differentiable function $f:\mathbb{R}^{I_1\times \dots \times I_N}\to \mathbb{R}$ is the gradient of the gradient and is defined for a tensor $A\in\mathbb{R}^{I_1\times \dots \times I_N}$ entry-wise by

@@

See @sec-hessian-from-gradient for how this definition can be reproduced by performing two gradients $\nabla^2 f = \nabla(\nabla f)$.

<!-- ::: {.callout-note}
We could have defined the Hessian of a scalar function as

$$
\nabla^2 f(A) [i_1,j_1, \dots, i_N, j_N] = \frac{\partial^2 f}{\partial A[i_1, \dots , i_N]\partial A[j_1, \dots , j_N]} (A).
$$

Notice how the dimensions are reordered to have all $i_n$'s followed by all the $j_n$'s rather than paired together in our definition. Pairing them together as shown in @eq-hessian often leads to cleaner results when using the usual outer product $\otimes$.
::: -->

::: {#def-frobenius-norm}

@@

For matrices $M$, the ($2\to 2$) operator norm is defined as

$$
\left\lVert M \right\rVert_{\text{op}} = \sup_{\lVert v \rVert_2 = 1} \left\lVert Mv \right\rVert_2 = \sigma_1(M)
$$

@@

For tensors $T$, the operator-norm is ambiguous since there are multiple ways we can treat tensors as function on other tensors. There is a canonical way to do this for vectors $x \mapsto v^{\top}x$ and matrices $x \mapsto Mx$, but not tensors. In the case of the Hessian tensor $\nabla^2 f(A)\in\mathbb{R}^{(I_1\times\dots\times I_N)^2}$ evaluated at $A\in\mathbb{R}^{I_1\times\dots\times I_N}$, it is natural to consider the function $X \mapsto \nabla^2 f(A) \times_{[N]} X$ for $X\in\mathbb{R}^{I_1\times\dots\times I_N}$. This gives us our definition of the operator norm on tensors.

::: {#def-operator-norm}

# Operator Norm

The operator norm of a half-symmetric tensor $A\in\mathbb{R}^{(I_1\times\dots\times I_N)^2}$ (@eq-half-symmetric) is defined as

$$
\left\lVert A \right\rVert_{\text{op}} = \sup_{\left\lVert X \right\rVert_F = 1} \left\lVert A \times_{[N]} X \right\rVert_F.
$$ {#eq-operator-norm}

In @eq-operator-norm, $X\in\mathbb{R}^{I_1\times\dots\times I_N}$. Note that this definition agrees with the usual operator norm on matrices when $N=1$.

::: {#thm-operator-norm-outer-product}

# Norm of an outer product

Let $T = A_1 \otimes \dots \otimes A_N \in \mathbb{R}^{(I_1\times \dots \times I_N)^2}$ where $A_n\in\mathbb{R}^{I_n\times I_n}$ are symmetric matrices.

Then $T$ is half-symmetric and

$$
\left\lVert T \right\rVert_{\text{op}} = \prod_{n=1}^N \left\lVert A_n \right\rVert_{\text{op}}.
$$

::: {.callout-warning}

According to how the outer product $\otimes$ is defined in @def-outer-product, the product $A_1 \otimes \dots \otimes A_N$ shown in @thm-operator-norm-outer-product is really an element of $\mathbb{R}^{I_1\times I_1 \times \dots \times I_N\times I_N}$. Note how the indexes are ordered differently than an element of $\mathbb{R}^{(I_1\times \dots \times I_N)^2} = \mathbb{R}^{I_1\times \dots \times I_N \times I_1\times \dots \times I_N}$. Correcting for this with explicit notation becomes cumbersome and would require tensor transposes, a new definition of an outer product, or reordering of indexes in the definition of a half-symmetric tensor. These can have knock-on effects to the definition of the Hessian, multi-mode product, and the operator norm.

To avoid the headache, the equality

$$
T=A_1 \otimes \dots \otimes A_N
$$

should be thought of as the following entry-wise equation

$$
T[i_1,\dots,i_N,j_1,\dots,j_N] = A_1[i_1,j_1] \cdots A_N[i_N,j_N].
$$ {#eq-corrected-outer-product}

With the outer product understood as @eq-corrected-outer-product, the results of @thm-operator-norm-outer-product that $T$ is half-symmetric and its operator norm is the product of the operator norms of the constituent matrices is true.

<!-- ^[This is relevant to @thm-quadratic-smoothness and @sec-lipschitz-computation where the Lipschitz step-size is computed in terms of the Operator norm of the Hessian of our objective function.] -->

<!-- = \sigma_1(\text{mat}(\nabla^2 f(A)))
where $\text{mat}:\mathbb{R}^{(I_1\times \dots \times I_N)^2}\to\mathbb{R}^{(I_1 \dots I_N)\times (I_1 \dots I_N)}$ takes a $2N$-ordered tensor and flattens it into a square $(I_1 \dots I_N)\times (I_1 \dots I_N)$ matrix. -->

::: {#def-lipschitz}

# Lipschitz Function

@@

::: {#def-smooth}

# Smooth Function

@@

Then:

@@

$$
L = \left\lVert D \right\rVert _{\text{op}}.
$$

@@

For the family of Tucker decompositions, we can compute the Lipschitz constants of the gradient efficiently similar to how we compute the gradient in @sec-gradient-computation with the following corollaries of @thm-quadratic-smoothness.

::: {#cor-least-squares-matrix}

Let $B\in\mathbb{R}^{R_1\times\dots\times R_N}$, $A_m\in\mathbb{R}^{I_m\times R_m}$, and $Y\in\mathbb{R}^{I_1\times\dots\times I_N}$. The function

$$
f(A)=\frac{1}{2} \left\lVert [B; A_1,\dots,A_{n-1},A,A_{n+1},\dots,A_N] - Y \right\rVert_F^2
$$

is quadratic, and $L$-smooth with constant

$$
L_{A_n} = \left\lVert \tilde{X}_n \cdot_n \tilde{X}_n \right\rVert _{\text{op}}
$$

::: {.proof}
The result follows from @thm-operator-norm-outer-product and @thm-quadratic-smoothness.
:::

::: {#cor-least-squares-core}

Let $A_n\in\mathbb{R}^{I_n\times R_n}$, and $Y\in\mathbb{R}^{I_1\times\dots\times I_N}$. The function

$$
f(B)=\frac{1}{2} \left\lVert [B; A_1,\dots,A_N] - Y \right\rVert_F^2
$$

is quadratic, and $L$-smooth with constant

$$
L_{B} = \prod_{n=1}^N\left\lVert A_n^\top A_n \right\rVert _{\text{op}}.
$$

::: {.proof}
The result follows from @thm-operator-norm-outer-product and @thm-quadratic-smoothness.
:::

This yields the following efficient implementations.

It is tempting to use the identity $\left\lVert A^\top A \right\rVert_{\text{op}} = \left\lVert A \right\rVert_{\text{op}}^2$ to calculate the Lipschitz constant without forming $A^\top A$. For tall dense matrices, using this identity is slower and more memory intensive as of Julia 1.11.2. See [LinearAlgebra.jl issue 1185](https://github.com/JuliaLang/LinearAlgebra.jl/issues/1185) on Github.

<!-- ::: callout-note
For the family of Tucker decompositions, the objective function $f$ is "block-quadratic" with respect to the factors. We saw in @eq-tucker-1-gradient-2 that the gradient is an affine map where $B$ is multiplied by the "slope" $A^\top A$ plus a shift of $-Y\times_1 A^\top$.

@@

TODO Explain how we get the following. General idea is to take partial Hessian of our quadratic function with respect to one block, and get that the operator norm of the hessian (as a map from $\mathbb{R}^{I_1\times \dots \times I_N}$ to itself) is the Lipschitz constant of the gradient. See Obsidian note "Gradients Hessians and Lipschitz Constants of a Tucker Decomposition". -->
```julia
function make_lipschitz(T::Tucker, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    N = ndims(T)
    if n==0 # the core is the zeroth factor
        function lipschitz_core(X::AbstractTucker; kwargs...)
            return prod(A -> opnorm(A'A), matrix_factors(X))
        end
        return lipschitz_core

    elseif n in 1:N # the matrix is the zeroth factor
        function lipschitz_matrix(X::AbstractTucker; kwargs...)
            B = core(X)
            matrices = matrix_factors(X)
            X̃ₙ = tuckerproduct(B, matrices; exclude=n)
            return opnorm(slicewise_dot(X̃ₙ, X̃ₙ; dims=n))
        end
        return lipschitz_matrix

@@

In the case of Tucker decomposition, the Lipschitz constants simplify to

$$
L_{B} = \left\lVert A^{\top}A\right\rVert_{\text{op}}, \quad L_{B} = \left\lVert B \cdot_{1} B\right\rVert_{\text{op}}.
$$

```julia
function make_lipschitz(T::Tucker1, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    if n==0 # the core is the zeroth factor
        function lipschitz0(X::Tucker1; kwargs...)
            A = matrix_factor(X, 1)
            return opnorm(A'A)
        end
        return lipschitz0

    elseif n==1 # the matrix is the zeroth factor
        function lipschitz1(X::Tucker1; kwargs...)
            B = core(X)
            return opnorm(slicewise_dot(B, B))
        end
        return lipschitz1

    else
        error("No $(n)th factor in Tucker1")
    end
end
```

Lastly, for CP decomposition, the Lipschitz constants for the matrices can be calculated similarly to the Tucker decomposition.

```julia
function make_lipschitz(T::CPDecomposition, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    N = ndims(T)
    if n in 1:N
        function lipschitz_matrix(X::CPDecomposition; kwargs...)
            id = core(X)
            matrices = matrix_factors(X)
            X̃ₙ = tuckerproduct(id, matrices; exclude=n)
            return opnorm(slicewise_dot(X̃ₙ, X̃ₙ; dims=n))
        end
        return lipschitz_matrix

    else
        error("No $(n)th factor in CPDecomposition")
    end
end
```

@@

In the case of Tucker decomposition, one modification of the update shown in @eq-proximal-explicit would be to update each column $a_{n,r}$, $r=1,\dots,R_n$ of the matrix $A_n$ separately. This would be suitable if the constraint that $A_N\in\mathcal{C}_n$ can be broken up further into the constraints $a_{n,r}\in\mathcal{C}_{n,r}$. This is shown in the following update scheme:

$$
a_{n,r}^{t+1} \leftarrow P_{\mathcal{C}}_{n,r}\left( a_{n,r}^{t} - \frac{1}{L_{n,r}^t}\nabla f_{n,r}^t(a_{n,r}^{t})\right),
$$ {#eq-sub-block-update-proper}

where $f_{n,r}^t(a)=\frac{1}{2} \left\lVert [B; A_1^{t+1},\dots,A_{n-1}^{t+1},A_{n,r}^t(a),A_{n+1}^{t},\dots,A_N^{t}] - Y \right\rVert_F^2$ and

$$
A_{n,r}^t(a) = \begin{bmatrix}
\uparrow &  & \uparrow & \uparrow & \uparrow &  & \uparrow \\
a_{n,1}^{t+1} & \cdots & a_{n,r-1}^{t+1}  & a  & a_{n,r+1}^{t}  & \cdots & a_{n,R_{n}}^{t}  \\
\downarrow &  & \downarrow & \downarrow & \downarrow &  & \downarrow \\
\end{bmatrix}.
$$ {#eq-proper-column-update}

In theory, the block update shown in @eq-sub-block-update-proper should be a bit more expensive than using the larger blocks on the matrices $A$ shown in @eq-proximal-explicit, since the gradient needs to be recomputed $R_n$ times for each matrix block $n$, rather than only computing the gradient once per block $n$. To get around this, we use the fact that $\nabla f_{n,r}^t(a)$ is the $r$th column from the gradient $\nabla f_{n}^t(A)$ where $f_{n}^t(A)=\frac{1}{2} \left\lVert [B; A_1^{t+1},\dots,A_{n-1}^{t+1},A,A_{n+1}^{t},\dots,A_N^{t}] - Y \right\rVert_F^2$. So we can approximate @eq-sub-block-update-proper by first calculating the gradient $\nabla f_{n}^t$ at

$$
\hat{A}_{n}^t = \begin{bmatrix}
\uparrow &  & \uparrow & \uparrow & \uparrow &  & \uparrow \\
a_{n,1}^{t} & \cdots & a_{n,r-1}^{t}  & a_{n,r}^{t}  & a_{n,r+1}^{t}  & \cdots & a_{n,R_{n}}^{t}  \\
\downarrow &  & \downarrow & \downarrow & \downarrow &  & \downarrow \\
\end{bmatrix},
$$ {#eq-merged-column-update}

and then updating each sub-block $r$ according to

$$
a_{n,r}^{t+1} \leftarrow P_{\mathcal{C}}_{n,r}\left( a_{n,r}^{t} - \frac{1}{L_{n,r}^t}\nabla f_{n}^t(\hat{A}_{n}^t)\right).
$$ {#eq-sub-block-update-half-merged}

Note the difference between @eq-proper-column-update and @eq-merged-column-update is that we don't use the most recent columns $a_{n,j}$ for $j<r$ in @eq-merged-column-update.

The update given in @eq-sub-block-update-half-merged can be merged back to an update on the whole block $A_n$

$$
A_{n}^{t+1} \leftarrow P_{\mathcal{C}}_{n}\left( A_{n}^{t} -\nabla f_{n}^t(A_{n}^t)(\hat{L}_{n}^t)^{-1} \right)
$$ {#eq-sub-block-update}

where we have the $R_n\times R_n$ diagonal "Lipschitz Matrix"

$$
\hat{L}_{n}^t = \begin{bmatrix}
{L}_{n,1}^t & 0 &  & 0 \\
0 & {L}_{n,2}^t &  & 0 \\
 &  & \ddots & \vdots \\
0 & 0 & \cdots & {L}_{n,R_{n}}^t
\end{bmatrix}.
$$

It is not too hard to show that ${L}_{n,r}^t$ is the Euclidean norm of the $r$th column^[We could have used the $r$th row of $\tilde{X}_n \cdot_n \tilde{X}_n$ since this matrix is symmetric. Since Julia store matrices in column-major order, many operations that perform column-wise are more efficient than their equivalent row-wise operation.] of the matrix $\tilde{X}_n \cdot_n \tilde{X}_n$ from @cor-least-squares-matrix,

$$
{L}_{n,r}^t = \left\lVert \left(\tilde{X}_n \cdot_n \tilde{X}_n \right)[:, r] \right\rVert _{2}.
$$

This leads to the following efficient calculation of the Lipschitz matrix in Julia.

```julia
function diagonal_lipschitz_matrix(T::Tucker, n::Int; kwargs...)
    B = core(T)
    matrices = matrix_factors(T)
    X̃ₙ = tuckerproduct(B, matrices; exclude=n)
    return Diagonal_col_norm(slicewise_dot(X̃ₙ, X̃ₙ; dims=n))
end

Diagonal_col_norm(X) = Diagonal(norm.(eachcol(X)))
```

We can now compare the merged sub-block update @eq-sub-block-update to the standard projected gradient descent update shown in @eq-proximal-explicit. The difference is that we calculate a "matrix step-size" $\hat{L}_{n}^t \in\mathbb{R}^{R_n \times R_n}$ rather than a scalar ${L}_{n}^t\in\mathbb{R}$. In practice, this leads to an improvement in convergence speed for two reasons.

First, computing the matrix $\hat{L}_{n}^t$ often faster than the scalar ${L}_{n}^t=\left\lVert \tilde{X}_n \cdot_n \tilde{X}_n \right\rVert _{\text{op}}$. The former only requires calculating the Euclidean norm of $R$ vectors for a total cost of $2R_{n}^2$ floating point operations (FLOPs), whereas the latter requires the top eigenvalue of $\tilde{X}_n \cdot_n \tilde{X}_n$. This is usually done with a power method or truncated SVD which can be costlier than the flat rate of $2R_{n}^2$ FLOPs.

Secondly, using the matrix $\hat{L}_{n}^t$ means columns where ${L}_{n,r}^t$ is small can take larger descent steps. This is because the largest singular value of $\tilde{X}_n \cdot_n \tilde{X}_n$ is an upper bound on the Euclidean norm of each column: ${L}_{n,r}^t \leq {L}_{n}^t$. Using the scaler Lipschitz ${L}_{n}^t$ is equivalent to the diagonal matrix

$$
D = \begin{bmatrix}
{L}_{n}^t & 0 &  & 0 \\
0 & {L}_{n}^t &  & 0 \\
 &  & \ddots & \vdots \\
0 & 0 & \cdots & {L}_{n}^t
\end{bmatrix}
$$

in the merged sub-block update shown in @eq-sub-block-update. So each column of $A_n$ is forced to use the worst case (largest) singular value of $\tilde{X}_n \cdot_n \tilde{X}_n$. In this way, the matrix $\hat{L}_{n}^t$ acts like a cheap approximate Hessian as if we were doing a quasi-Newton update with step-size $1$.

For completeness, we can perform the same merged sub-block update to update the core $B$. In this case, we obtain the more complicated "Lipschitz tensor" $\hat{L}_{B}^t \in\mathbb{R}^{(R_1\times\dots\times R_N)^2}$ defined by

$$
\hat{L}_{B}^t = \hat{L}_{B,1}^t \otimes \dots \otimes \hat{L}_{B,N}^t
$$

where each matrix $\hat{L}_{B,n}^t \in\mathbb{R}^{R_n\times R_n}$ is diagonal with non-zero entries

$$
{L}_{B,n}^t [r,r] = \left\lVert \left((A_n^t)^\top A_n^t\right)[:, r] \right\rVert _{2}.
$$

The merged sub-block update for the core becomes

$$
B^{t+1} \leftarrow P_{\mathcal{C}}_{B}\left( B^{t} - \nabla f_{0}^t(B^t) \times_B (\hat{L}_{B}^t)^{-1} \right)
$$ {#eq-sub-block-update-core}

with the multiplication

$$
\begin{align}
\nabla f_{0}^t(B^t) \times_B (\hat{L}_{B}^t)^{-1} &= \nabla f_{0}^t(B^t)  \bigtimes_{ n} (\hat{L}_{B,n}^{t})^{-1} = \llbracket \nabla f_{0}^t(B^t) ; (\hat{L}_{B,1}^{t})^{-1},\dots,(\hat{L}_{B,n}^{t})^{-1} \rrbracket.
\end{align}
$$

This should be thought of as normalizing each dimension of the tensor $\nabla f_{0}^t(B^t)$ so that we can take a unit step-size.

Putting the core and matrices Lipschitz calculations together gives us the following Julia code. Note we store $\hat{L}_{B}^t$ in factored form as a tuple of diagonal matrices to save space and computation.

```julia
function make_block_lipschitz(T::AbstractTucker, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    N = ndims(T)
    if n==0 # the core is the zeroth factor
        function lipschitz_core(X::AbstractTucker; kwargs...)
            return map(A -> Diagonal_col_norm(A'A), matrix_factors(X))
        end # Returns a tuple of diagonal matrices
        return lipschitz_core

    elseif n in 1:N
        function lipschitz_matrix(X::AbstractTucker; kwargs...)
            matrices = matrix_factors(X)
            X̃ₙ = tuckerproduct(core(X), matrices; exclude=n)
            return Diagonal_col_norm(slicewise_dot(X̃ₙ, X̃ₙ; dims=n))
        end
        return lipschitz_matrix

@@

In practice, we find that extrapolating the iterate based on the prior iterate

$$
\hat{A}_n^t \leftarrow A_n^t + \omega_n^t (A_n^{t} - A_n^{t-1})
$$

for some amount of extrapolation $\omega_n^t \geq 0$ before applying the update @eq-sub-block-update greatly improves the speed of descent. This can be thought of as a type of momentum where we continue to move in directions that showed a lot of improvement during the last iteration.

Our selection for $\omega_n^t$ follows Xu and Yin's method for block coordinate descent @xu_BlockCoordinateDescent_2013, which is itself inspired by Tseng and Yun's coordinate gradient descent method @tseng_coordinate_2009.

TODO CONTINUE FROM HERE

```julia
function (U::MomentumUpdate)(x::T; x_last::T, ω, δ, kwargs...) where T
    n = U.n
    if checkfrozen(x, n)
        return x
    end
    # TODO avoid redoing this lipschitz calculation and instead store the previous L
    # TODO generalize this momentum update to allow for other decaying momentums ω
    L = U.lipschitz(x; kwargs...)
    L_last = U.lipschitz(x_last; kwargs...)
    ω = min.(ω, δ .* .√(L_last/L))

    a, a_last = factor(x, n), factor(x_last, n)

    # Equivalent mathematically, but slightly less efficient ways of applying momentum
    # @. a = a + ω * (a - a_last)

    # @. a += ω * (a - a_last)

    # a .*= 1 + a
    # a .-= ω .* a_last

    a = U.combine(a, id + ω) # handle diagonal Lipschitz constants
    a .-= U.combine(a_last, ω)
end
```

```julia
"""update parameters needed for the update"""
function initialize_parameters(decomposition, Y, previous; momentum::Bool, random_order, recursive_random_order, kwargs...)
	# parameters for the update step are symbol => value pairs
	# they are held in a dictionary since we may mutate these for ex. the stepsize
	parameters = Dict{Symbol, Any}()

	# General Looping
	parameters[:iteration] = 0
	parameters[:x_last] = previous[begin] # Last iterate
	parameters[:random_order] = random_order
	parameters[:recursive_random_order] = recursive_random_order

	# Momentum
	if momentum
		parameters[:t_last] = float(1) # need this field to hold Floats, not Ints
		parameters[:t] = update_t(float(1))
		parameters[:ω] = (parameters[:t_last] - 1) / parameters[:t]
		parameters[:δ] = kwargs[:δ]
	end

	function updateparameters!(parameters, decomposition, previous)
		parameters[:iteration] += 1
		#parameters[:x_last] = previous[begin] # Last iterate updated with updateprevious!
											   # this ensures we only copy over the decomposition
											   # once since updateparameters! happens before updateprevious!

											   # This is commented since parameters[:x_last] already points to previous[begin]
		# Momentum
		if momentum
			parameters[:t_last] = parameters[:t]
			parameters[:t] = update_t(parameters[:t_last])
			parameters[:ω] = (parameters[:t_last] - 1) / parameters[:t]
		end
	end

	return parameters, updateparameters!
end
```

@@

# Appendix {.appendix}

## Building the Hessian from two gradients {#sec-hessian-from-gradient}

To build the Hessian from the definition of the gradient, we first extend the gradient to tensor-valued functions. For a function $F:\mathbb{R}^{J_{1}\times \dots \times J_{N}}\to\mathbb{R}^{I_{1}\times\dots \times I_{M}}$ where

$$
F(X)=[f_{i_{1}\dots i_{M}}(X)]
$$

is a tensor of scalar functions $f_{i_{1}\dots i_{M}}:T\to \mathbb{R}$, the gradient of $F$ at $X$ is defined entry-wise as

$$
\begin{align}
\nabla F(X)[i_{1},\dots,i_{M}][j_{1},\dots,j_{N}]&=\nabla f_{i_{1}\dots i_{M}}(X)[j_{1},\dots,j_{N}] \\
&=\frac{{\partial f_{i_{1}\dots i_{M}}}}{\partial X[j_{1},\dots,j_{N}]} (X).
\end{align}
$$

This treats the gradient at $X$ as a tensor of tensors $\nabla F(X)\in (\mathbb{R}^{I_{1}\times\dots \times I_{M}})^{J_{1}\times \dots \times J_{N}}$. This is naturally isomorphic to a tensor of order $M+N$ with entries

$$
\nabla F(X)[i_{1},\dots,i_{M},j_{1},\dots,j_{N}]=\frac{{\partial f_{i_{1}\dots i_{M}}}}{\partial X[j_{1},\dots,j_{N}]} (X).
$$ {#eq-tensor-gradient-entries}

So we conclude that the gradient at $X$ of a tensor-valued function $F$ is $\nabla F: \mathbb{R}^{J_{1}\times \dots \times J_{N}}\to \mathbb{R}^{I_{1}\times\dots \times I_{M}\times J_{1}\times \dots \times J_{N}}$ is given by @eq-tensor-gradient-entries.

We can define the Hessian of a scalar function $f:\mathbb{R}^{I_{1}\times\dots \times I_{N}}\to \mathbb{R}$ at $X$ as $\nabla^2 f(X) = \nabla (\nabla f) (X):\mathbb{R}^{I_{1}\times\dots \times I_{N}}\to \mathbb{R}^{(I_{1}\times\dots \times I_{N})^2}$. The inner nabla $\nabla$ is the gradient of the scalar function $f$, and the outer nabla $\nabla$ is the gradient of the tensor-valued function $\nabla f$.

This means

$$
\nabla^2 f(A) [i_1, \dots, i_N, j_1, \dots , j_N] = \frac{\partial^2 f}{\partial A[j_1, \dots , j_N]\partial A[i_1, \dots , i_N]} (A),
$$

but if the function has continuous second derivatives, we can perform the partial derivatives in either order

$$
\frac{\partial^2 f}{\partial A[j_1, \dots , j_N]\partial A[i_1, \dots , i_N]} (A) = \frac{\partial^2 f}{\partial A[i_1, \dots , i_N]\partial A[j_1, \dots , j_N]} (A).
$$
